Datasheet for 2018 Canadian General Social Survey on Giving, Volunteering & Participating

### 2022-04-28 inputs/data/datasheet.Rmd
added datasheet

---
title: "The Qualities of Volunteering that predict higher Life Satisfaction: A multiple linear regression model"
subtitle: "Insights from the 2018 Canadian General Social Survey"
author: Kimlin Chin
thanks: "Code and data are available at: https://github.com/KCtt457/gssvolunteering2018."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Volunteering has been associated with higher life satisfaction; however, there is limited research on the precise aspects of volunteering associated with life satisfaction. In this paper, we investigate how different aspects of volunteering including number of organizations, frequency and different qualities of experience in volunteering are correlated with life satisfaction using a multiple linear regression model. We find that among people who volunteer, believing that you made a contribution to the community and having been meaningfully involved in the work of the organization were correlated with higher life satisfaction. Our findings have implications for the general public in making the most out of their volunteer experience. \\par \\textbf{Keywords:} volunteering, life satisfaction, multiple linear regression, canadian general social survey, canada"
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = '') # remove NAs from kable https://stackoverflow.com/questions/27626461/hiding-nas-when-printing-a-dataframe-in-knitr

library(tidyverse)
library(kableExtra)
library(patchwork)

gss1 <- read_csv(here::here("inputs/data/gss.csv"))
```

# Introduction

<!-- Background, context -->
Life satisfaction is an indicator of well-being, and although it has many different definitions, I like the one by Sumner (1966) as cited in @lifesatisdef, which is "A positive  evaluation  of  the conditions of your life, a judgment that at least on balance, it measures up favorably against your standards or expectations." Volunteering, which is giving of your time and effort freely towards helping others outside of your household without compensation, has been positively associated with life satisfaction [@Thoits2001]. However, previous studies have mainly investigated the effect of volunteer status alone and not the different aspects of volunteering beyond number of hours or frequency [@BORGONOVI20082321; @socsci7050078]. Many studies also prefer to examine older populations [@socsci7050078; @HUANG2019172]. The few studies that have investigated the aspects of volunteering associated with well-being, however, have found that being appreciated for their volunteer work and believing that others benefit from their efforts [@Jongenelis2021a], volunteering 2-3 hours per week [@morrow], organizational support [@geront], volunteering for causes they believe to be important, and meaningful volunteer activities [@Jongenelis2021] were positive predictors of well-being.

In this paper we analyze data from the 2018 Canadian General Social Survey on Giving, Volunteering & Participating to investigate the effects of volunteering and the different aspects of volunteering on life satisfaction. Specifically we will compare how different forms of volunteering (formal vs informal) as well as other altruistic behavior such as giving financially are correlated with life satisfaction. We construct a multiple linear regression model to do this. We also create another multiple linear regression model where life satisfaction is explained by certain aspects of volunteering including number of organizations, frequency, and different qualities of the volunteer experience. We find that while formal volunteering, informal volunteering and giving were all positively correlated with life satisfaction, formal volunteering had the strongest effect. Our second model revealed that among people who volunteer, believing you made a contribution to the community and being involved in meaningful ways in the work of the organization were significant aspects of volunteering that predicted higher life satisfaction. This adds to the literature since we investigated the population in Canada whereas previous studies looked at other parts of the world (US [@geront], Australia [@Jongenelis2021], Europe [@socsci7050078], Asia [@HUANG2019172]), we compared different types of volunteering whereas other studies usually chose one type (formal) and we looked at other aspects of volunteering not commonly explored before: improvement of job opportunities and use of skills and experience.

The rest of the paper has the following structure: Section \@ref(data) describes the data from the 2018 Canadian General Social Survey on Giving, Volunteering & Participating, Section \@ref(model) discusses the multiple linear regression models used in the data analysis, Section \@ref(results) presents the results of our analysis and Section \@ref(discussion) discusses the findings and limitations. `R` [@citeR] and the R packages `tidyverse`[@thereferencecanbewhatever], `kableExtra` [@kableExtra] and `patchwork` [@patchwork] were used for data processing and to make the plots and tables in this paper.

# Data

## Data Source

The data is from the 2018 Canadian General Social Survey (GSS) on Giving, Volunteering & Participating. Canada's GSS program conducts independent cross-sectional surveys each year on a specific topic. The main objectives of the program are to gather data on social trends in order to monitor changes in the living conditions and well-being of Canadians and to provide information on specific social policy issues. The topic for 2018 was Giving, Volunteering & Participating, which was the seventh time data on this topic has been collected at the national level.

## Data Collection and Methodology

<!-- Population, frame and sample -->
Data was collected by Statistics Canada during the period of September 4th to December 28th, 2018. The target population included individuals 15 years and over in Canada, excluding institutionalized persons and residents of the Yukon, Northwest Territories, and Nunavut. For sampling, the ten provinces were divided into strata by geographical area. The survey frame was developed by combining lists of telephone numbers from various sources and lists of dwellings from the Address Register. Each record in the survey frame was assigned a stratum within its province. Households were then randomly sampled within each stratum and a single respondent from each household was chosen using an age-order selection method to either complete an electronic questionnaire or telephone interview.

Due to the low prevalence of volunteers in the population, rejective sampling was also used to obtain a larger number of respondents in the population of interest which is volunteers. There were 16,149 respondents in the survey, excluding 'rejected' respondents.

## Key Features

There are 956 variables and 16,149 observations in the dataset.

The variables include basic demographic information, such as age, gender, marital status, province, as well as volunteer specifics and details, reasons for volunteering or not volunteering, quality and history of volunteering, financial giving and youth experiences.

A subset of the variables is shown in table \@ref(tab:datafeatures).

```{r datafeatures, echo=FALSE, message=FALSE}
gss1 %>%
  select(dh1gage, gndr, fv1fvol, vs1_010, vs1_020, rv_090, qvs_030) %>%
  slice(1:10) %>%
  kable(
    caption = "Some key features",
    col.names = c("Age group of respondent", "Gender", "Volunteer Flag", "Number of Organizations", "Volunteer Frequency", "Reasons for Volunteering-To use skills", "Quality of Volunteer Experience-Community Contribution"),
    digits = 1,
    booktabs = TRUE,
    linesep = ""
  ) %>%
column_spec(c(1,4,5,6,7), width = "2.5cm")
```

Since we are interested in the relationship between volunteering and life satisfaction, we conduct data exploration of these features. Figure \@ref(fig:volunteerflag) shows the counts of formal volunteers, informal volunteers and givers compared to their counterparts in the sample. Formal volunteers are defined as individuals who worked within a volunteer organization in the past 12 months whereas informal volunteers are not members of an organization but engaged in activities that helped people or the community directly during the past 12 months. Givers are respondents who made a donation to charitable or non-profit organization within the past 12 months.

```{r volunteerflag, echo=FALSE, fig.cap="Number of Volunteers (Formal and Informal) and Givers in the Sample", message=FALSE}
fvol <- ggplot(gss1) +
  geom_bar(aes(x = fv1fvol, fill=fv1fvol)) +
  labs(x="Formal Volunteer Flag",
       y="") +
  scale_y_continuous(limits = c(0, 12500), breaks = seq(0, 12500, by=4000)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(),
        legend.position = "none")

ivol_data <- gss1 %>%
  mutate(informalvol = factor(ivc_flag, levels = c("Not an informal volunteer", "Informal volunteer")),
         giver = factor(fg1fgiv, levels = c("Non-giver", "Giver")))

ivol <- ggplot(ivol_data) +
  geom_bar(aes(x = informalvol, fill=informalvol)) +
  labs(x="Informal Volunteer Flag",
       y="Number of responses") +
  scale_y_continuous(limits = c(0, 12500), breaks = seq(0, 12500, by=4000)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(),
        legend.position = "none")

giving <- ggplot(ivol_data) +
  geom_bar(aes(x = giver, fill=giver)) +
  labs(x="Gave financially",
       y="") +
  scale_y_continuous(limits = c(0, 12500), breaks = seq(0, 12500, by=4000)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(),
        legend.position = "none")

fvol / ivol / giving
```

There is about a 50-50 split between the formal volunteers and non-volunteers in our sample. For informal volunteers and givers, the ratio is about the same with 3 informal volunteers/givers to 1 not informal volunteer/non-giver.  However, this should be looked at with caution when making generalizations to the wider population given that the data was obtained via rejective sampling.

Figure \@ref(fig:lifesatis) shows life satisfaction by volunteer status. Life satisfaction was rated on a scale from 0 to 10, where 0 means very dissatisfied with life, and 10 means very satisfied with life. Overall, the majority of respondents had relatively high satisfaction ratings in the range 7-10. It can clearly be seen from the graph that in the lower life satisfaction ratings 0-6, the percentage of non-volunteers was greater than volunteers. Interestingly at a rating of 7 it appears to be about an even split, and for higher life satisfaction ratings the percentage of volunteers is greater than the percentage of non-volunteers.

```{r lifesatis, echo=FALSE, message=FALSE, fig.cap="Life satisfaction by Volunteer Status"}

lifesat <- gss1 %>%
  select(slm_01, fv1fvol) %>%
  mutate(slm_01 = as.factor(slm_01)) %>%
  na.omit()

ggplot(lifesat) +
  geom_bar(aes(x=slm_01, fill=fv1fvol), position = "dodge") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank()) +
  labs(x = "Life Satisfaction",
       y = "Number of Responses",
       fill = "Volunteer Flag",
       caption = "0 = Very dissatisfied,\n10 = Very satisfied")

```

Therefore this brief glimpse of data hints at a relationship between being a volunteer and life satisfaction.

We would also like to explore the characteristics of volunteering that may be involved in this relationship. We will restrict our sample to volunteers only when exploring the characteristics of volunteering, since it is not applicable to non-volunteers. It now begs the question, which aspects of volunteering to investigate? The data includes dozens of variables on the types of organizations and types of volunteer activities respondents are involved in. The large number of variables would thus make it difficult to fit to a concise statistical model, and since a person may be involved in multiple organizations and activities, it may also be difficult to measure the individual effects of certain activities.

Some features we could consider instead include number of organizations and frequency of volunteering, which are shown in figure \@ref(fig:freqorg). We also look at different benefits experienced from volunteering, which are the variables labelled 'Quality of the Volunteering Experience' in the dataset. The different 'qualities' which are rated on a 5-degree scale from Strongly disagree to Strongly agree include networking, job opportunities, community contribution, health improved, meaningful involvement and use of skills and experience. We convert these to binary variables with 'Yes' if they agreed to experiencing the benefit and 'No' otherwise for conciseness. Figure \@ref(fig:conciseqove) shows the number of 'Yes' instances to the different qualities.

```{r freqorg, echo=FALSE, fig.cap="Frequency and Number of Organizations volunteered with in the past 12 months"}

# Add collapsed binary variables for Quality of Volunteering Experience
gss1 <- gss1 %>%
  mutate(lifesatisfaction = slm_01, numorgs = vs1_010, freq = vs1_020, 
         network = ifelse(qvs_010 %in% c("Agree", "Strongly agree"), "Yes", "No"),
         jobops = ifelse(qvs_020 %in% c("Agree", "Strongly agree"), "Yes", "No"),
         community = ifelse(qvs_030 %in% c("Agree", "Strongly agree"), "Yes", "No"), 
         health = ifelse(qvs_040 %in% c("Agree", "Strongly agree"), "Yes", "No"), 
         meaningful = ifelse(qvs_050 %in% c("Agree", "Strongly agree"), "Yes", "No"),
         skills = ifelse(qvs_060 %in% c("Agree", "Strongly agree"), "Yes", "No"))

volunteer_data <- gss1 %>%
  filter(fv1fvol == "Volunteer") %>%
  select(numorgs, freq) %>%
  na.omit() %>%
  mutate(freq_ord = str_replace(freq, "three or four", "3-4")) %>%
  mutate(freq_ord = str_replace(freq_ord, "in ", "in\n")) %>%
  mutate(freq_ord = factor(freq_ord, levels = c("Daily or almost daily", "At least once a week", "At least once a month", "At least 3-4 times in\nthe past 12 months", "Once or twice in\nthe past 12 months")))

freq_graph <- ggplot(volunteer_data) +
  geom_bar(aes(x = freq_ord)) +
   theme_minimal() +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor.y = element_blank()) +
  labs(title = "A. Frequency of\n    Volunteering",
    x = "",
       y = "Number of responses",
       fill = "") +
  coord_flip()

numorgs_graph <- ggplot(volunteer_data) +
  geom_bar(aes(x = numorgs)) +
   theme_minimal() +
  theme(panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank()) +
  labs(title = "B. Number of\n    Organizations",
       x = "Number of organizations",
       y = "Number of responses",
       fill = "")

freq_graph + numorgs_graph
```

```{r conciseqove, echo=FALSE, fig.cap="Qualities of the Volunteering Experience that respondents agreed to experiencing"}

qove_bars <- gss1 %>%
  filter(fv1fvol == "Volunteer") %>%
  select(network, jobops, community, health, meaningful, skills) %>%
  pivot_longer(everything(), names_to = "quality", values_to = "Yes") %>%
  filter(Yes == "Yes") %>%
  group_by(quality) %>%
  summarise(count = n())

qove_description <- tibble(description = c("Networking", "Job opportunities",
                       "Community contribution", "Improve health/well-being",
                       "Meaningful Involvement", "Use skills and experience"),
       quality = c('network', 'jobops', 'community','health','meaningful', 'skills'))

qove <- left_join(qove_bars, qove_description, by = "quality")

ggplot(qove) +
  geom_bar(aes(x = reorder(description, count), y = count), stat = "identity", fill = "#00bfc4") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor.y = element_blank()) +
  labs(x = "Quality of Volunteering",
       y = "Number of responses",
       fill = "") +
  coord_flip()
```

We also considered including the reasons people choose to volunteer and if they get what they wanted out of their volunteer experience, that is, if their volunteer experience was fulfilling. This can be done by matching the stated reasons for volunteering with the variables for the quality of their experience (e.g. reason 'to network' is matched with the quality 'networking'). However, we found that the reasons were almost always fulfilled by the volunteer experience for the majority of participants, so we excluded this from our model analysis since it does not add much additional information given the benefits themselves.

```{r concisefulfilling, include=FALSE}

# Add new variables for fulfilling
gss1 <- gss1 %>%
  mutate(fulfillingn = case_when(rv_040 == "Yes" & qvs_010 %in% c("Agree", "Strongly agree") ~ "Yes",
                                 TRUE ~ "No"),
         fulfillingj = case_when(rv_050 == "Yes" & qvs_020 %in% c("Agree", "Strongly agree") ~ "Yes",
                                 TRUE ~ "No"),
         fulfillingcc = case_when(rv_080 == "Yes" & qvs_030 %in% c("Agree", "Strongly agree") ~ "Yes",
                                 TRUE ~ "No"),
         fulfillings = case_when(rv_090 == "Yes" & qvs_060 %in% c("Agree", "Strongly agree") ~ "Yes",
                                 TRUE ~ "No"),
         fulfillingh = case_when(rv_110 == "Yes" & qvs_040 %in% c("Agree", "Strongly agree") ~ "Yes",
                                 TRUE ~ "No"),
         fulfilling = case_when(rv_040 == "Yes" & qvs_010 %in% c("Agree", "Strongly agree") ~ "Yes",
                                rv_050 == "Yes" & qvs_020 %in% c("Agree", "Strongly agree") ~ "Yes",
                                rv_080 == "Yes" & qvs_030 %in% c("Agree", "Strongly agree") ~ "Yes",
                                rv_090 == "Yes" & qvs_060 %in% c("Agree", "Strongly agree") ~ "Yes",
                                rv_110 == "Yes" & qvs_040 %in% c("Agree", "Strongly agree") ~ "Yes",
                                TRUE ~ "No"))
```

```{r fulfilplot, include=FALSE, fig.cap="Reasons for Volunteering that were fulfilled"}

reasons_v <- gss1 %>%
  filter(fv1fvol == "Volunteer") %>%
  select(rv_040, rv_050,
         rv_080, rv_090, rv_110) %>%
  pivot_longer(everything(), names_to = "Reason", values_to = "Yes") %>%
  filter(Yes == "Yes") %>%
  group_by(Reason) %>%
  summarise(count = n())

fulfilling_v <- gss1 %>%
  filter(fv1fvol == "Volunteer") %>%
  select(fulfillingn, fulfillingj, fulfillingcc, fulfillings, fulfillingh) %>%
  pivot_longer(everything(), names_to = "fulfilled", values_to = "Yes") %>%
  filter(Yes == "Yes") %>%
  group_by(fulfilled) %>%
  summarise(fcount = n())

reasons_described <- tibble(Reason = c('rv_040', 'rv_050',
                  'rv_080', 'rv_090', 'rv_110'),
       description = c("To network", "Job opportunities",
                       "Community contribution", "To use your skills",
                       "Improve health/well-being"),
       fulfilled = c('fulfillingn', 'fulfillingj', 'fulfillingcc','fulfillings','fulfillingh'))

reasons_v <- left_join(reasons_v, reasons_described, by = "Reason")
reasons_v <- left_join(reasons_v, fulfilling_v, by = "fulfilled")

colors <- c("Fulfilled" = "#00bfc4", "Not Fulfilled" = "#f8766d")

reasons_plot <- ggplot(reasons_v) +
  geom_bar(aes(x = reorder(description, count), y = count), stat = "identity",fill = "#f8766d") +
  geom_bar(aes(x = reorder(description, count), y = fcount), stat = "identity", fill = "#00bfc4") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor.y = element_blank()) +
  labs(title = "Reasons for Volunteering in the past 12 months",
       x = "",
       y = "Number of responses",
       fill = "Fulfilled") +
  scale_fill_manual(values = colors) +
  coord_flip()

volunteersonly <- gss1 %>%
  filter(fv1fvol == "Volunteer")

fulfilling_plot <- ggplot(volunteersonly) +
  geom_bar(aes(x=fulfilling, fill = fulfilling)) +
  theme_minimal() + 
  theme(panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank()) +
  labs(title = "  B. At least one reason\n      was fulfilled",
       x = "",
       y = "Number of responses",
       fill = "")

reasons_plot
```

## Strengths and Weaknesses

Some strengths of the data include that it is a fairly large sample and it explores multiple different variables on the topic of interest which is giving, volunteering and participating in detail. Additionally, although volunteers were the subject of interest, data on non-volunteers were also included such as their reasons for not volunteering. I did not encounter any weaknesses in the data, except perhaps that we cannot make generalizations to the wider population of Canada about the proportion of individuals who volunteer since rejective sampling was used.

# Model

## Model 1: Life Satisfaction and Volunteer Flag

For our first model, we are interested in how volunteer status is correlated with life satisfaction. We construct a multiple linear regression model with life satisfaction as the dependent variable, and the main independent variables of interest are formal volunteer flag, informal volunteer flag and giver flag. We include informal volunteer and giver in our analysis as points of comparison to formal volunteering. The other dependent variables in our model are various socio-demographic variables including gender, age, education level, marital status, if there are children at home, employment status and family income as these have been found to be relevant explanatory variables for life satisfaction in previous research [@Thoits2001; @BORGONOVI20082321]. The model is show in \@ref(eq:model1) follows:

<!-- https://tex.stackexchange.com/questions/112238/how-to-wrap-a-long-equation-in-latex -->
\begin{equation}
\begin{aligned}
life\_satisfaction = \beta_1 formal\_volunteer + \beta_2 informal\_volunteer + \beta_3 giver + \beta_4 gender + \beta_5 age \\ + \beta_6 education + \beta_7 married + \beta_8 children\_at\_home + \beta_9 employment\_status + \beta_{10} income\_level + \epsilon (\#eq:model1)
\end{aligned}
\end{equation}

The $\beta$'s are the coefficients of our independent variables whose values will be determined when we fit the model, and $\epsilon$ is the random error term.

```{r model1r, include=FALSE, message=FALSE}

model1var <- gss1 %>%
  select(lifesatisfaction,
    fv1fvol,
         gndr,
         dh1gage,
         marstat,
         cxrflag,
         dlfs,
         famincg2,
         dh1ged,
    fg1fgiv,
    ivc_flag) %>%
  mutate(
    isvolunteer = fv1fvol,
    agegroup = factor(dh1gage, levels = c("25-34 years", "15-24 years", "35-44 years", "45-54 years", "55-64 years", "65 years and over")),
    gender = factor(gndr, levels = c("Male", "Female")),
    married = ifelse(marstat == "Married", "Yes", "No"),
    childrenathome = cxrflag,
    employed = dlfs,
    familyincome = factor(
      famincg2,
      levels = c(
        "$50,000 to $74,999",
        "Less than $25,000",
        "$25,000 to $49,999",
        "$75,000 to $99,999",
        "$100,000 to $124,999",
        "$125,000 and more"
      )
    ),
    education = dh1ged,
    giver = factor(fg1fgiv, levels = c("Non-giver", "Giver")),
    informal_volunteer = factor(ivc_flag, levels = c("Not an informal volunteer","Informal volunteer"))
  ) %>%
  na.omit()

model1 <- lm(lifesatisfaction ~ isvolunteer + informal_volunteer + giver + gender + agegroup + education + married + childrenathome + employed + familyincome, data = model1var)

summary(model1)

```

## Model 2: Life Satisfaction and Volunteer Characteristics

For our second model, we construct a multiple linear regression model to investigate what specific characteristics of formal volunteering might be correlated with life satisfaction. For this model, we restrict the datapoints to formal volunteers only (since non-volunteers do not have volunteer characteristics). Our dependent variable is life satisfaction again, and the independent variables include number of volunteer organizations, volunteer frequency and different benefits of volunteering experienced including networking, improvement in job opportunities, community contribution, meaningful involvement, improved health and use of skills and experience. The model is shown in \@ref(eq:model2) follows:

\begin{equation}
\begin{aligned}
life\_satisfaction = \beta_1 num\_organizations + \beta_2 frequency + \beta_3 network + \beta_4 jobops \\ + \beta_5 community + \beta_6 meaningful + \beta_7 health + \beta_8 skills + \epsilon  (\#eq:model2)
\end{aligned}
\end{equation}

The $\beta$'s are the coefficients of our independent variables whose values will be determined when we fit the model, and $\epsilon$ is the random error term.

```{r model2r, include=FALSE, message=FALSE}

model2var <- gss1 %>%
  filter(fv1fvol == "Volunteer") %>%
  select(lifesatisfaction, numorgs, freq, fulfilling, fulfillingn, fulfillings, fulfillingh, fulfillingj, fulfillingcc, network, jobops, community, health, meaningful, skills) %>%
  na.omit()

# model2var %>%
#   group_by(fulfilling) %>%
#   summarise(count = n())

# model2 <- lm(lifesatisfaction ~ numorgs + freq + fulfilling + network + jobops + community + health + meaningful + skills, data = model2var)

model2 <- lm(lifesatisfaction ~ numorgs + freq + network + jobops + community + health + meaningful + skills, data = model2var)

summary(model2)

# model2 <- lm(lifesatisfaction ~ numorgs + freq + fulfillingn + fulfillings + fulfillingh + fulfillingj + fulfillingcc + network + jobops + community + health + meaningful + skills, data = model2var)
# 
# summary(model2)

```

# Results

Table \@ref(tab:tab1) shows the results of our first model with formal volunteer status, informal volunteer status, giver and the socio-demographic variables. Estimates of the coefficients of our dependent variables are shown in the 'Estimate' column, and their corresponding p-values are shown in the 'p-value' column. P-value is defined as "the probability under the assumption of no effect or no difference (null hypothesis), of obtaining a result equal to or more extreme than what was actually observed" [@Dahiru2011]. Since it is a probability, the interpretation of p-values close to zero is that the result is unlikely due to chance, which we will call "significant". For our analysis, we will assume a significance level of 0.05, that is, if the p-value is less than 0.05 we will call the variable significant.

The baseline of 7.19 represents the life satisfaction score of male, unmarried and employed Canadians who do not volunteer formally or informally, do not give financially, are in the age range 25-34 years, have a highschool degree, no children at home and have a family income in the range \$50,000 to \$74,999. Recall that life satisfaction is rated on a scale of 0-10, so this is already on the higher side of life satisfaction. Significant variables that were associated with an increase in life satisfaction in descending order of largest coefficient include being married, 65 years or older, formal volunteering, earning an income of \$125,000 and more, giving financially, being female and being an informal volunteer. Being married vs not married from the baseline resulted in an increase of 0.41 points in life satisfaction and being a formal volunteer increased life satisfaction by 0.30 points.

Significant variables associated with a decrease in life satisfaction include being not employed, having a family income of less than $50,000, having less than a highschool degree, being 35-54 years of age, and having children at home. Being unemployed vs employed resulted in a decrease of 0.84 points from the baseline whereas having children at home only had a small negative effect of 0.08 points decrease in life satisfaction.

```{r tab1, echo=FALSE}
ests <- format(round((summary(model1)$coeff)[,1], 2), nsmall = 2)
pvals <- format(signif((summary(model1)$coeff)[,c( 4)], digits = 3), nsmall = 2)
rownames_for_table <- c("Baseline", "Volunteer", "Female", "Age Group 15-24 years", "Age Group 35-44 years", "Age Group 45-54 years", "Age Group 55-64 years", "Age Group 65 years and over", "Education Less than High School", "Education Post-secondary diploma", "Education University Diploma", "Married Yes", "Children at home Yes", "Employed Not in labour force", "Employed Unable to determine", "Employed Unemployed", "Family income Less than $25,000", "Family income $25,000 to $49,999", "Family income $75,000 to $99,999", "Family income $100,000 to $124,999", "Family income $125,000 and more")
rownames_for_table1 <- c("Baseline Life Satisfaction Score", "Formal Volunteer", "Informal Volunteer", "Giver", "Female", "Age Group", "", "", "", "", "Education", "", "", "Married", "Children at home", "Employment Status", "", "", "Family income", "", "", "", "")
v_subcategories <- c("", "", "", "", "", "15-24 years", "35-44 years", "45-54 years", "55-64 years", "65 years and over", "Less than High School", "Post-secondary diploma", "University Diploma", "", "", "Not in labour force", "Unable to determine", "Unemployed", "Less than $25,000", "$25,000 to $49,999", "$75,000 to $99,999", "$100,000 to $124,999", "$125,000 and more")
colnames_for_table <- c("","Estimate", "p-value*")
my_pretty_table <- cbind(v_subcategories, ests, pvals)
rownames(my_pretty_table) <- rownames_for_table1
colnames(my_pretty_table) <- colnames_for_table
knitr::kable(my_pretty_table, align = c("l","r", "r"), caption = "Estimated Coefficients for predictors of Model 1", booktabs = TRUE,
    linesep = "") %>%
  kable_styling(latex_options = "hold_position") %>%
  add_footnote(c("p-value is the probability under the assumption of no effect or no difference (null hypothesis), of obtaining a result equal to or more extreme than what was actually observed (Dahiru 2011). If p-value < 0.05, we say the variable is significant."), notation = "symbol")
```

Table \@ref(tab:tab2) shows the results of our second model which considers the characteristics of volunteering. The baseline here represents the life satisfaction score of individuals that volunteer at least once a month, and experienced none of the listed 5 benefits of volunteering. The significant variables include contributing to the community, being meaningfully involved in the work of the organization and increasing number of organizations. Making a community contribution has an estimated 0.26 point increase in life satisfaction score and being meaningfully involved in the organization has a 0.29 increase in life satisfaction score from the baseline. Each additional organization that a person volunteered with only results in a small increase of 0.05 points in life satisfaction. Significant variables that are negatively correlated with life satisfaction include having experienced an improvement in job opportunities because of the volunteer experience and volunteering at least 3-4 times in the past 12 months but less than once a month. Having benefited from improved job opportunities and volunteering at least 3-4 times in the past 12 months were associated with 0.22 point decrease and 0.20 decrease respectively.

```{r tab2, echo=FALSE}
ests <- format(round((summary(model2)$coeff)[,1], 2), nsmall = 2)
pvals <- format(signif((summary(model2)$coeff)[,c( 4)], digits = 3), nsmall = 2)
rownames_for_table1 <- c("Baseline Life Satisfaction Score", "Number of Organizations", "Frequency", "", "", "", "Networking", "Job opportunities", "Community Contribution", "Improved Health", "Meaningful Involvement", "Use Skills and Experience")
v_subcategories <- c("", "", "At least once a week", "At least three or four times in the past 12 months", "Daily or almost daily", "Once or twice in the past 12 months", "", "", "", "", "", "")
colnames_for_table <- c("","Estimate", "p-value")
my_pretty_table <- cbind(v_subcategories, ests, pvals)
rownames(my_pretty_table) <- rownames_for_table1
colnames(my_pretty_table) <- colnames_for_table
knitr::kable(my_pretty_table, align = c("l","r", "r"), caption = "Estimated Coefficients for predictors of Model 2", booktabs = TRUE,
    linesep = "") %>%
  kable_styling(latex_options = "hold_position")
```

# Discussion

<!-- brief summary of paper -->
The present study investigated the effect of volunteering on life satisfaction, with three main research questions in mind: 1) Are volunteers more satisfied with life than non-volunteers?, 2) How do different types of altruistic behaviour (formal volunteering, informal volunteering and giving) affect life satisfaction? and 3) Among volunteers, what specific characteristics of formal volunteering are associated with higher life satisfaction? We created two multiple linear regression model using data from Canada's 2018 General Social Survey on Giving, Volunteering & Participating to address these questions. The results showed that compared to non-volunteers, the life satisfaction score of volunteers was higher, so the answer to the first research question is yes. Formal volunteering, informal volunteering and giving were all positively correlated with life satisfaction, but formal volunteering had the strongest effect. To address the third question, we considered the number of organizations volunteered with, frequency of volunteering, and various benefits obtained from the volunteer experience such as improved job opportunities and community contribution. The results showed that being meaningfully involved in the volunteer organization and contributing to the community were positively correlated with life satisfaction, whereas obtaining job opportunities was negatively correlated with life satisfaction. This suggests that volunteering that concerns itself with the growth and development of the community gives more life satisfaction than volunteering for one's own personal benefit. This has implications for making the most out of your volunteer experience.

## Volunteering increases life satisfaction

From our first model, we saw that being a volunteer (both formal and informal) was significant in predicting higher life satisfaction even after taking into account the effects of various socio-demographic variables known to be correlated with life satisfaction. In fact, formal volunteering had the third strongest effect after marriage and being 65 years and older. The positive correlation of volunteering with life satisfaction aligns with previous studies [@Thoits2001].

Informal volunteering also had a positive effect on life satisfaction but it was 0.17 points lower than for formal volunteering. This result contradicts the result of the study by @Appau2018 which found life satisfaction to be higher for informal volunteering than formal volunteering using a very similar model to ours. In their study, the authors used construal-level theory from Psychology to hypothesize that informal volunteering would lead to higeher subjective well-being since the persons could "relate better with the beneficiaries of the volunteer work compared to formal volunteering which tends to be more structured and regulated" and that informal volunteer work is "community based and more interpersonal" [@Appau2018].

The different results in our study could perhaps be explained by the difference in culture (since their study was UK-based, as opposed to our study which uses Canadian data) and experience in formal volunteering. In our data, two of the most common qualities of experience in formal volunteering that were agreed upon by most  volunteers in the sample were that the individual felt that they have been involved in meaningful ways in the work of the organization and that they have made a contribution to the community. Unfortunately data was not collected on the qualities of informal volunteer experience so we cannot make a direct comparison, but we can see that formal volunteers get the "community based and more interpersonal" experience that @Appau2018 describes for informal volunteering over formal volunteering for persons in the UK. This experience may not be the same for informal volunteering in Canada, and so we see the higher life satisfaction for formal volunteers over informal volunteers. Further research into the experiences of informal and formal volunteers is needed to make stronger conclusions and the existing literature is still quite limited (@Appau2018 was one of the few studies I could find on this topic).

Giving donations also had a significant positive effect on life satisfaction as well, which was higher than informal volunteering and close to formal volunteering, only about 0.05 points lower. A previous study using US data by @choi_kim_2011 found that any amount of charitable donations had a higher positive effect on well-being than volunteering; however, the difference was small. Thus, the result is similar to our study in that there is a small difference between the effects of giving and volunteering on life satisfaction so we can conclude that making donations and volunteering produce similar increases in life satisfaction.

So in general it appears that any form of helping others results in higher life satisfaction, but out of the different types, formal volunteering gives the highest increase in life satisfaction. So what is it about formal volunteering that increases satisfaction? We discuss this in the analysis of our second model.

## Community and Meaningful involvement in Volunteering: the secret to life satisfaction?

From the second model, having a meaningful involvement in the volunteer organization and contributing to the community were the significant predictors of higher life satisfaction among volunteers. On the other hand, experiencing improved job opportunities as a result of volunteering was associated with a decrease in life satisfaction. Use of skills and experience, networking and improved health and well-being were not significant predictors compared to the other variables.

This hints at the idea that volunteering should be a selfless act you do in groups if you want to derive life satisfaction from it, or perhaps it is just that altruistic people become volunteers to engage in these types of activities and hence get more life satisfaction since they are doing what they enjoy. Our model only gives a correlation so we cannot make any causal claims. This is an interesting idea to consider given that in Western cultures the idea of self-actualization is often promoted, which is realizing one's full potential, so it is natural to think that developing yourself should make you have a more satisfied life. However, life also needs balance, too much of anything is not good, so the proportion of selflessness demonstrated through volunteering may only be one piece of the cake that makes life satisfactory.

Coupling this with insights from our first model, there is just something about the selfless organizational community contributions of formal volunteering that leads to higher life satisfaction that you don't get through informally helping others or financial donations. More research into this theme can investigate the amount of value persons put on their own self-development vs group/community development when volunteering and how this various across different personalities and cultures, as this would influence what aspects of life they derive satisfaction from. It is also important to keep in mind that life satisfaction varies from person to person and that these conclusions we draw are from trends in the general Canadian population.

## Weaknesses and next steps

A limitation of this study is that many of the variables used in the analysis are subjective responses of the participants so they may not provide the most accurate true measure of certain variables. For example, there was only one question on life satisfaction in the survey where participants rated for themselves their life satisfaction on a scale of 0-10. Although many previous studies use a similar measure when investigating this topic [@Thoits2001; @Appau2018], more robust measures of life satisfaction exist such as the Satisfaction with Life Scale (SWLS) [@swls]. Another limitation is that we did not go into depth on the effects of the aspects of informal volunteering on life satisfaction, as our focus was formal volunteering and also comparable features on informal volunteering were not provided in the survey data. For next steps, more data can be collected on the aspects of informal volunteering to see how they are correlated with life satisfaction.

\appendix

# Appendix {-}

## Datasheet

>>>>

---
title: "Datasheet for 2018 Canadian General Social Survey on Giving, Volunteering & Participating"
author: Kimlin Chin
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references_datasheet.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Extract of the questions from @gebru2021datasheets
